% ============================================================================
% Acknowledgments Page
% ============================================================================

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

We would like to express our sincere gratitude to all those who contributed to the successful completion of this graduation project.

First and foremost, we extend our deepest appreciation to our supervisor, \textbf{Dr. Diaa Emad}, for their invaluable guidance, continuous support, and constructive feedback throughout this project. Their expertise and encouragement have been instrumental in shaping our work.

We are also grateful to the \textbf{Mechatronics Department} at the Faculty of Engineering, Ain Shams University, for providing us with the necessary resources, facilities, and academic environment to pursue this project.

Special thanks to the \textbf{technical staff} and \textbf{laboratory personnel} who assisted us with equipment, testing facilities, and technical support.

We would also like to thank our \textbf{colleagues and friends} who provided moral support, shared their knowledge, and contributed to valuable discussions throughout this journey.

Finally, we are forever indebted to our \textbf{families} for their unconditional love, patience, and encouragement. Their support has been the foundation of our success.

\vspace{1cm}
\begin{flushright}
\textit{The AUSRA Team}\\
Cairo, 2026
\end{flushright}
